Local sensitivity analysis of a supercritical extraction model

### 2025-06-07 CJCE-24-1047.R2.tex
07.06

	% ===================================================
	% METADATA
	% ===================================================
	\title[mode=title]{Local Sensitivity Analysis of a Supercritical Extraction Model}                      
	\shorttitle{Sensitivity Analysis}

@@

	\begin{keywords}
		Supercritical extraction \sep Sensitivity analysis \sep Mathematical modelling
	\end{keywords}

@@

	One of the most popular applications of supercritical CO$_2$ is the extraction of essential oils, as described by many researchers, for example, by \citet{Sodeifian2017}, \citet{Reverchon1993} or \citet{Sovova1994}. Traditional methods, such as distillation and organic solvent extraction, are commonly employed but have drawbacks. Distillation involves high temperatures that can lead to the thermal degradation of heat-sensitive compounds. This limitation has increased the popularity of alternative techniques, such as supercritical fluid extraction. Supercritical CO$_2$ is appealing due to its distinctive properties: it is inflammable, non-toxic and non-corrosive. Supercritical fluids can exhibit both gas- and liquid-like properties, allowing for adjustable dissolving power through changes in operating conditions.

	The applications of supercritical carbon dioxide are not limited only to an extraction process but can also be used for impregnation, as described by \citet{Weidner2018} or \citet{Machado2022}. Impregnation is defined as modifying the properties of bulk substances by physically or chemically binding/adsorbing impregnates to a bulk material or surface, such as the hydrophobization of surfaces. The main advantage of using supercritical CO$_2$ is that, after depressurization, it desorbs from the surface and evaporates, leaving a solvent-free product. On the other hand, the main disadvantage of using carbon dioxide for impregnation is the low solubility of many drugs of interest.
	The study of \citet{Ameri2020} investigates the loading of lansoprazole into polymers using supercritical carbon dioxide and examines how various parameters, such as temperature, pressure, and time, affect the drug loading efficiency. The results indicate that increasing any of these parameters enhances drug loading, with temperature having the most significant impact. \citet{Fathi2022} explored the use of supercritical carbon dioxide to enhance the bioavailability of ketoconazole by impregnating it into water-soluble polymers, specifically polyvinylpyrrolidone and hydroxypropyl methylcellulose. Utilizing a Box-Behnken design, the researchers optimized the impregnation process by varying pressure, temperature, and time, achieving increase in drug loadings ranging.

	Another application of supercritical CO$_2$ is nanoparticle formation, as investigated by \citet{Padrela2018}, \citet{Franco2021} or \citet{Sodeifian2022}. Supercritical carbon-dioxide-assisted technologies enable the production of different morphologies of different sizes, including nanoparticles and nanocrystals, by modulating the operating conditions. Supercritical fluid-based processes have advantages over techniques conventionally employed to produce nanosized particles or crystals, such as reduced use of toxic solvents. Moreover, the CO$_2$ is removed from the final product by simple depressurization. 
	\citet{Sodeifian2018} investigated the solubility of Letrozole, a poorly water-soluble anticancer drug, in supercritical carbon dioxide with and without menthol as a solid co-solvent. The addition of menthol increased Letrozole's solubility by 7.1 times compared to supercritical CO$_2$ alone. Using the rapid expansion of supercritical solutions with solid co-solvent method, the average particle size of Letrozole was reduced to the nanoscale. Temperature was found to have the most significant impact on nanoparticle size reduction, while pressure had the least effect. 
	The study of \citet{SaadatiArdestani2020} explored the preparation of phthalocyanine green nano pigment using supercritical carbon dioxide as an antisolvent. The researchers employed the gas antisolvent technique to achieve nano-sized particles of the pigment.
	\citet{Sodeifian2019} analysed the production of amiodarone hydrochloride nanoparticles using an ultrasonic-assisted rapid expansion of supercritical solution into a liquid solvent method. By optimizing parameters such as pressure, temperature, and polymeric stabilizer concentration, the researchers achieved significant particle size reduction. Characterization techniques confirmed the successful formation of nanoparticles with improved properties.

	This study investigates the extraction of essential oil from chamomile flowers (Matricaria chamomilla L.) via supercritical fluid extraction techniques and the modelling of this process. Chamomile is a medicinal herb widely cultivated in southern and eastern Europe — in countries such as Germany, Hungary, France and Russia. It can also be found outside Europe, for instance in Brazil, as discussed by \citet{Singh2011}. This plant is distinguished by its hollow, bright gold cones, housing disc or tubular florets and surrounded by about fifteen white ray or ligulate florets. Chamomile has been used for its medicinal benefits, serving as an anti-inflammatory, antioxidant, mild astringent, and healing remedy. Extracts of chamomile are widely used to calm nerves and mitigate anxiety, hysteria, nightmares, insomnia and other sleep-related conditions, according to \citet{Srivastava2009}. \citet{Orav2010} reported that oil yields from dried chamomile samples ranged from 0.7 to 6.7 mL/kg. The highest yields of essential oil, between 6.1 and 6.7 mL/kg, were derived from chamomile sourced from Latvia and Ukraine. In comparison, chamomile from Armenia exhibited a lower oil content of 0.7 mL/kg.

@@

	Depending on the needs one of two approaches can be considered while developing a mathematical model for the extraction process. A model, which is based on a multiple regression can be used if the relation between inputs and outputs is the only of interests. \citet{Sodeifian2017a} investigated an influence of pressure, temperature and particle size on the extraction efficiency of oil from dracocephalum kotschyi boiss seed. The second order polynomial model was applied to obtain the corresponding response surface and to identify the optimum operating conditions. 
	The study of \citet{Sodeifian2017b} investigates the of essential oil from Eryngium billardieri, focusing on optimizing extraction conditions and developing a mathematical model based on the second order polynomial to predict the process yield. The researchers employed a simulated annealing algorithm to optimize parameters such as pressure, temperature, and extraction time, aiming to maximize oil efficiency.

	Alternatively, a first principle model can be derived and applied to cover not only the input-output relations, but the phenomena occurring in the system. This approach allows for more detailed representation the system behaviour, but requires deeper understanding of the underlying physics and more rigorous experiments.

@@

	This work builds upon the linear kinetic model suggested by \citet{Reverchon1996}, deriving fundamental governing equations to develop a comprehensive model for the chamomile oil extraction process. This model aims at control-oriented simplicity, assuming semi-continuous operation within a cylindrical vessel. The process involves a supercritical solvent being pumped through a fixed bed of finely chopped biomass to extract the solute, followed by separation of the solvent and solute in a flush drum to collect the extract. Parameters such as pressure ($P$), feed flow rate ($F$) and inlet temperature ($T_{in}$) are adjustable and measurable, while the outlet temperature ($T_{out}$) and the amount of product at the outlet can only be monitored. Figure \ref{fig: SFE_drawing} presents a simplified process flow diagram.

@@

	The primary aim of this study is to analyse the influence of changes in operating conditions on the process model developed by \citet{Sliczniuk2024}. Understanding how variations in parameters and control variables affect the model's behaviour is essential for improving its accuracy, robustness, and practical applicability. To achieve this, a comprehensive sensitivity analysis is employed, serving as a systematic approach to assess the impact of model parameters and controls on the outputs. The results of this sensitivity analysis are valuable for multiple purposes. First, they help identify sources of uncertainty within the model, highlighting which parameters most significantly influence predictions. Second, they provide insights for potential model simplification, allowing the exclusion of non-essential parameters without compromising accuracy. Finally, the analysis can detect inconsistencies or errors by uncovering unexpected relationships between inputs and outputs, which may signal flaws in the model's assumptions or structure.

@@

	Different supercritical extraction models have been analysed using various sensitivity analysis techniques in the literature. For instance, \citet{Fiori_2007} performed sensitivity calculations by varying parameters within their confidence intervals and observing the changes in model results. Their analysis revealed that the particle diameter and internal mass transfer coefficient significantly influence extraction during the diffusion-control regime.

	\citet{Santos2000} considered the model of \citet{Sovova1994} for semi-continuous isothermal and isobaric extraction processes using carbon dioxide as a solvent. They conducted a parametric sensitivity analysis using a two-level factorial design, disturbing model parameters by 10\% and analysing their main effects. They proposed strategies for high-performance operation based on sensitivities related to superficial velocity, particle diameter, initial solute concentration in the solid phase and solute concentration in the fluid phase at the extractor inlet.

@@

	In a his work, \citet{Turanyi1990} presented a comprehensive review of sensitivity analysis methods for spatially homogeneous, constant-parameter chemical kinetic systems. This study explored a range of sensitivity analysis techniques, including concentration sensitivity, rate sensitivity, and feature sensitivity analysis. Practical applications of these methods in chemical kinetics were highlighted, showcasing their utility in understanding and predicting system behavior.

@@

	In their comprehensive review, \citet{Saltelli2005} examined sensitivity analysis methods applicable to chemical reaction modeling. They discussed both local and global techniques, emphasizing the role of global sensitivity analysis in understanding complex chemical systems. The authors highlighted the importance of variance-based methods in quantifying the influence of input parameters on model outputs by discussing results of a few simulation analysis.

	A key novelty of this work lies in its rigorous application of derivative-based local sensitivity analysis to a first-principles model of supercritical fluid extraction. While multiple publications available in the literature discuss application of sensitivity analysis on a supercritical extraction model, many of these either omit local sensitivity analyses and focus on basic parametric comparisons (\citet{Santos2000}, \citet{Fiori_2007} or \citet{Zahedi2010}). In contrast, the present approach integrates a quasi-1D partial-differential-equation model with automatic differentiation of the mass and energy balances, enabling a detailed exploration of how infinitesimal changes in parameters, which propagates propagate through the system. As the result, corresponding system response of the state space (with emphasis on solutes concentrations on solid and liquid phase as well as yield). Although this work focus on the influence of a few selected parameters, the same methodology can be extended to any other parameter. Notably, this level of detail is crucial near the critical region of CO$_2$, where small deviations in process conditions can trigger large changes in solvent properties (such as density and solubility). This approach thus provides a systematic and quantitative measure of the system’s sensitivity to parameter fluctuations across different extraction regimes, from kinetic- to diffusion-controlled.

@@

	The set of quasi-one-dimensional governing equations in Cartesian coordinates is described by Equations \ref{EQ: CompressibleEuler_1} - \ref{EQ: CompressibleEuler_3}:

@@

	where $\rho_f$ is the density of the fluid, $A_f$ is the function which describes a change in the cross-section, $v$ is the velocity, $P$ is the total pressure, $e$ is the internal energy of the fluid, $t$ is time and $z$ is the spatial direction.

@@

	Assuming that the mass flow rate is constant in time, the temporal derivative becomes the mass flux F, and the spatial derivative can be integrated along $z$ as

@@

	To simplify the system dynamics, it is assumed that $F$ is a control variable and affects the whole system instantaneously (due to $\nabla \cdot u = 0$), which allows finding the velocity profile that satisfies mass continuity based on $F$, $\phi$ and $\rho_f$:

@@

	The fluid density $\rho_f$ can be obtained from the Peng-Robinson equation of state if the temperature and thermodynamic pressure are known along $z$. Variation in fluid density may occur due to pressure or inlet temperature changes. In a non-isothermal case, in Equations \ref{EQ:Velocity_1} and \ref{EQ:Velocity_2}, $\rho_f$ is considered the average fluid density along the extraction column.

@@

	As the solvent flows through the fixed bed, CO$_2$ molecules diffuse into the pores, adsorb on the inner surface and form a film due to solvent-solid matrix interactions. The dissolved solute diffuses into the bulk from the particle's core through the solid-fluid interface, the pore and the film. Figure \ref{fig: SFE_Mechanism} shows the mass transfer mechanism, where the mean solute concentration in the solid phase is denoted by $c_s$, and the equilibrium concentrations at the solid-fluid interface are denoted by $c_s^*$ and $c_p^*$ for the solid and fluid phases, respectively. The concentration of the solutes in the fluid phase in the centre of the pore is denoted by $c_p$. As the solute diffuses through the pore, its concentration changes, reaching $c_{pf}$ at the opening. Then, the solute diffuses through the film around the particle and reaches bulk concentration $c_f$. The two-film theory describes the solid-fluid interface inside the pore. The overall mass transfer coefficient can be determined from the relationship between the solute concentration in one phase and its equilibrium concentration.

@@

	where $\mu$ is sphericity, $l$ a characteristic dimension of particles that can be defined as $l = r/3$, $r$ is the mean particle radius, $\rho_s$ is the solid density, $D_i$ corresponds to the overall diffusion coefficient and $c_P^*$ is the concentration at the solid-fluid interface (which, according to the internal resistance model, is supposed to be at equilibrium with the fluid phase).

@@

	The volumetric partition coefficient $k_p$ acts as an equilibrium constant between the solute concentration in one phase and the corresponding equilibrium concentration at the solid-fluid interphase. \citet{Spiro2007} propose defining the mass partition coefficient $k_m$ as

@@

	According to \citet{Reverchon1996}, the kinetic term becomes

@@

	The choice between the two formulations should consider the system's properties and the underlying mass transfer assumptions. In systems where external mass transfer dominates, the resistance to solute transport primarily occurs at the fluid-solid interface, typically governed by the diffusion of solute across the fluid boundary layer. Under these conditions, the Bulley formulation is more appropriate, as it explicitly emphasizes the fluid-phase concentration ($c_f$) as the driving force for mass transfer through the term $c_f - c_p^*(c_s)$. Conversely, when internal mass transfer dominates, the resistance to solute transport is controlled by diffusion or adsorption processes within the solid phase. In such cases, the Reverchon formulation is more suitable, as it accounts for deviations in the equilibrium through the term $c_s - c_s^*(c_f)$.

@@

	Following the idea of the broken-and-intact cell (BIC) model (\citet{Sovova2017}), the internal diffusion coefficient $D_i$ is considered to be a product of the reference value of $D_i^R$ and the exponential decay function $\gamma$, as given by Equation \ref{EQ: C_sat_function}:

@@

	The empirical correlations for $D_i$ and $\Upsilon$ were derived by \citet{Sliczniuk2024} and validated for temperatures of $30 - 40~^\circ C$, pressures of $100 - 200$ bar, and mass flow rates of $3.33-6.67 \cdot 10^{-5}$ kg/s. Figures \ref{fig:Correlation_Di} and \ref{fig:Correlation_Gamma} show the results of multiple linear regression applied to solutions of parameter estimation and selected independent variables. The region marked with the white dashed line represents the confidence region, where the model has been tested. Both correlations should be equal or greater than zero to avoid unphysical behaviour such as  reverse mass transfer. The multiple linear regression functions are combined with the rectifier function to ensure non-negativity.

@@

	If the value of enthalpy $h$ is known from the time evolution of the energy equation, and pressure $P$ is known from measurement, then temperature $T$ can be reconstructed based on the departure function. The departure function is a mathematical function that characterizes the deviation of a thermodynamic property (enthalpy, entropy or internal energy) of a real substance from that of an ideal gas at the same temperature and pressure. The enthalpy departure function, as presented by \citet{Gmehling2019} for the Peng-Robinson equation of state,  is defined by Equation \ref{EQ:Enthalpy_PR}:

@@

	where $\alpha$ is defined as $\left( 1+\kappa \left( 1 - \sqrt{T_r} \right) \right)^2$, $T_r$ is the reduced temperature, $P_r$ is the reduced pressure, $Z$ is the compressibility factor, $\kappa$ is a quadratic function of the acentric factor and $B$ is calculated as $0.07780\frac{P_r}{T_r}$.

@@

	The method of lines is used to transform the process model equations into a set of ODEs denoted by $G(x;\Theta)$. For a derivative to be conservative, it must form a telescoping series. In other words, only the boundary terms should remain after adding all terms coming from the discretization over a grid, and the artificial interior points should be cancelled out. Discretization is applied to the conservative form of the model to ensure mass conservation. The backward finite difference is used to approximate the first-order derivative, while the central difference scheme approximates the second-order derivative $z$ direction. The length of the fixed bed is divided into $N_z$, i.e. equally distributed points in the $z$ direction. The state-space model after discretization is denoted by $x$ and defined as follows the Figure \ref{fig:discretization}, where $x \in \mathbb{R}^{N_x = 3N_z+2} $ and $\Theta \in \mathbb{R}^{N_\Theta =  N_{\theta} + N_u } $, $N_{\theta}$ is the number of parameters and $N_{u}$ is the number of control variables.

@@

	The sensitivity matrix $S(x;\Theta)$ represents how the state vector $x$ changes in response to variations in the model parameters $\Theta$. In physical terms, this means it measures how sensitive the system is to changes in the underlying parameters that define its behavior. The new system of equations can be obtained by taking the total derivative of $S$ with respect to time $t$:

@@

	This equation decomposes the total effect of parameter changes on the system dynamics into two parts. The indirect effect represented by the term involving $\bar{J}_x (x; \Theta)$ This term shows how changes in the parameters influence the system through their effect on the state variables. This is an indirect feedback effect where parameter changes alter the state, which in turn affects the system response. The direct effect given by $\bar{J}_x (x; \Theta)$ reflects the immediate sensitivity of the system to parameter variations.

@@

	The augmented system combines both the state equations and sensitivity equations into a unified system. This allows the simultaneous computation of both the system behavior and its sensitivity to parameter changes.

	The initial conditions are described as

@@

	Details of the process model and its parameters can be found in \citet{Sliczniuk2024}. The model was validated under the following operating conditions: temperatures from $30$ to $40 ^\circ C$, pressures from $100$ to $200$ bar, and mass flow rates from $3.33$ to $6.67 \cdot 10^{-5}$ kg/s. In this study, the local sensitivity analysis is used to investigate how changes in model parameters affect the state space and extraction yield, where the results reflect the system’s response to infinitesimal deviations in parameter values. Although this work focuses on two specific cases, the methodology can be extended to any model parameter. The first case assesses the empirical correlations embedded in the process model, specifically the zeroth-order term $\Upsilon(0)$ in the multiple linear regression for $\Upsilon$. The second case examines the effect of operating conditions, particularly pressure, on the state space and model outputs. For the first analysis, the midpoint of the validated range ($35 ^\circ C$, $150$ bar and $5\cdot 10^{-5}$) was selected. The second analysis is carried out in two steps. First, we examine the impact of pressure changes on the state variables (solute concentrations in both solid and fluid phases) at the same midpoint conditions. Second, we perform a yield analysis at the same temperature and mass flow rate, but at different pressures (100, 125, 150, 175, and 200 bar).

>>>>

	% ===================================================
	% METADATA
	% ===================================================
	\title[mode=title]{Local sensitivity analysis of a supercritical extraction model}                      
	\shorttitle{Sensitivity analysis}

@@

	\begin{keywords}
		supercritical extraction \sep sensitivity analysis \sep mathematical modelling
	\end{keywords}

@@

	One of the most popular applications of supercritical CO$_2$ is the extraction of essential oils, as described by many researchers, for example, by \citet{Sodeifian2017}, \citet{Reverchon1993}, or \citet{Sovova1994}. Traditional methods, such as distillation and organic solvent extraction, are commonly employed but have drawbacks. Distillation involves high temperatures that can lead to the thermal degradation of heat-sensitive compounds. This limitation has increased the popularity of alternative techniques, such as supercritical fluid extraction. Supercritical CO$_2$ is appealing due to its distinctive properties: it is inflammable, non-toxic, and non-corrosive. Supercritical fluids can exhibit both gas- and liquid-like properties, allowing for adjustable dissolving power through changes in operating conditions.

	The applications of supercritical carbon dioxide are not limited only to an extraction process but can also be used for impregnation, as described by \citet{Weidner2018} and \citet{Machado2022} Impregnation is defined as modifying the properties of bulk substances by physically or chemically binding/adsorbing impregnates to a bulk material or surface, such as the hydrophobization of surfaces. The main advantage of using supercritical CO$_2$ is that, after depressurization, it desorbs from the surface and evaporates, leaving a solvent-free product. On the other hand, the main disadvantage of using carbon dioxide for impregnation is the low solubility of many drugs of interest.
	The study by \citet{Ameri2020} investigates the loading of lansoprazole into polymers using supercritical carbon dioxide and examines how various parameters, such as temperature, pressure, and time, affect the drug loading efficiency. The results indicate that increasing any of these parameters enhances drug loading, with temperature having the most significant impact. \citet{Fathi2022} explored the use of supercritical carbon dioxide to enhance the bioavailability of ketoconazole by impregnating it into water-soluble polymers, specifically polyvinylpyrrolidone and hydroxypropyl methylcellulose. Utilizing a Box--Behnken design, the researchers optimized the impregnation process by varying pressure, temperature, and time, achieving increase in drug loadings ranging.

	Another application of supercritical CO$_2$ is nanoparticle formation, as investigated by \citet{Padrela2018}, \citet{Franco2021}, and \citet{Sodeifian2022} Supercritical carbon dioxide-assisted technologies enable the production of different morphologies of different sizes, including nanoparticles and nanocrystals, by modulating the operating conditions. Supercritical fluid-based processes have advantages over techniques conventionally employed to produce nanosized particles or crystals, such as reduced use of toxic solvents. Moreover, the CO$_2$ is removed from the final product by simple depressurization. 
	\citet{Sodeifian2018} investigated the solubility of Letrozole, a poorly water-soluble anticancer drug, in supercritical carbon dioxide with and without menthol as a solid co-solvent. The addition of menthol increased Letrozole's solubility by 7.1 times compared to supercritical CO$_2$ alone. Using the rapid expansion of supercritical solutions with solid co-solvent method, the average particle size of Letrozole was reduced to the nanoscale. Temperature was found to have the most significant impact on nanoparticle size reduction, while pressure had the least effect. 
	The study of \citet{SaadatiArdestani2020} explored the preparation of phthalocyanine green nano pigment using supercritical carbon dioxide as an antisolvent. The researchers employed the gas antisolvent technique to achieve nano-sized particles of the pigment.
	\citet{Sodeifian2019} analyzed the production of amiodarone hydrochloride nanoparticles using an ultrasonic-assisted rapid expansion of supercritical solution into a liquid solvent method. By optimizing parameters such as pressure, temperature, and polymeric stabilizer concentration, the researchers achieved significant particle size reduction. Characterization techniques confirmed the successful formation of nanoparticles with improved properties.

	This study investigates the extraction of essential oil from chamomile flowers (Matricaria chamomilla L.) via supercritical fluid extraction techniques and the modelling of this process. Chamomile is a medicinal herb widely cultivated in southern and eastern Europe---in countries such as Germany, Hungary, France and Russia. It can also be found outside Europe, for instance in Brazil, as discussed by \citet{Singh2011} This plant is distinguished by its hollow, bright gold cones, housing disc or tubular florets and surrounded by about fifteen white ray or ligulate florets. Chamomile has been used for its medicinal benefits, serving as an anti-inflammatory, antioxidant, mild astringent, and healing remedy. Extracts of chamomile are widely used to calm nerves and mitigate anxiety, hysteria, nightmares, insomnia and other sleep-related conditions, according to \citet{Srivastava2009}. \citet{Orav2010} reported that oil yields from dried chamomile samples ranged from 0.7 to 6.7 mL/kg. The highest yields of essential oil, between 6.1 and 6.7 mL/kg, were derived from chamomile sourced from Latvia and Ukraine. In comparison, chamomile from Armenia exhibited a lower oil content of 0.7 mL/kg.

@@

	Depending on the needs, one of two approaches can be considered while developing a mathematical model for the extraction process. A model which is based on a multiple regression can be used if the relation between inputs and outputs is the only one of interest. \citet{Sodeifian2017a} investigated the influence of pressure, temperature, and particle size on the extraction efficiency of oil from dracocephalum kotschyi boiss seed. The second order polynomial model was applied to obtain the corresponding response surface and to identify the optimum operating conditions. 
	The study of \citet{Sodeifian2017b} investigates the of essential oil from Eryngium billardieri, focusing on optimizing extraction conditions and developing a mathematical model based on the second order polynomial to predict the process yield. The researchers employed a simulated annealing algorithm to optimize parameters such as pressure, temperature, and extraction time, aiming to maximize oil efficiency.

	Alternatively, a first principle model can be derived and applied to cover not only the input--output relations, but the phenomena occurring in the system. This approach allows for more detailed representation the system behaviour, but requires deeper understanding of the underlying physics and more rigorous experiments.

@@

	This work builds upon the linear kinetic model suggested by \citet{Reverchon1996}, deriving fundamental governing equations to develop a comprehensive model for the chamomile oil extraction process. This model aims at control-oriented simplicity, assuming semi-continuous operation within a cylindrical vessel. The process involves a supercritical solvent being pumped through a fixed bed of finely chopped biomass to extract the solute, followed by separation of the solvent and solute in a flush drum to collect the extract. Parameters such as pressure ($P$), feed flow rate ($F$), and inlet temperature ($T_{in}$) are adjustable and measurable, while the outlet temperature ($T_{out}$) and the amount of product at the outlet can only be monitored. Figure \ref{fig: SFE_drawing} presents a simplified process flow diagram.

@@

	The primary aim of this study is to analyze the influence of changes in operating conditions on the process model developed by \citet{Sliczniuk2024}. Understanding how variations in parameters and control variables affect the model's behaviour is essential for improving its accuracy, robustness, and practical applicability. To achieve this, a comprehensive sensitivity analysis is employed, serving as a systematic approach to assess the impact of model parameters and controls on the outputs. The results of this sensitivity analysis are valuable for multiple purposes. First, they help identify sources of uncertainty within the model, highlighting which parameters most significantly influence predictions. Second, they provide insights for potential model simplification, allowing the exclusion of non-essential parameters without compromising accuracy. Finally, the analysis can detect inconsistencies or errors by uncovering unexpected relationships between inputs and outputs, which may signal flaws in the model's assumptions or structure.

@@

	Different supercritical extraction models have been analyzed using various sensitivity analysis techniques in the literature. For instance, \citet{Fiori2007} performed sensitivity calculations by varying parameters within their confidence intervals and observing the changes in model results. Their analysis revealed that the particle diameter and internal mass transfer coefficient significantly influence extraction during the diffusion-control regime.

	\citet{Santos2000} considered the model of \citet{Sovova1994} for semi-continuous isothermal and isobaric extraction processes using carbon dioxide as a solvent. They conducted a parametric sensitivity analysis using a two-level factorial design, disturbing model parameters by 10\% and analyzing their main effects. They proposed strategies for high-performance operation based on sensitivities related to superficial velocity, particle diameter, initial solute concentration in the solid phase, and solute concentration in the fluid phase at the extractor inlet.

@@

	In another work, \citet{Turanyi1990} presented a comprehensive review of sensitivity analysis methods for spatially homogeneous, constant-parameter chemical kinetic systems. This study explored a range of sensitivity analysis techniques, including concentration sensitivity, rate sensitivity, and feature sensitivity analysis. Practical applications of these methods in chemical kinetics were highlighted, showcasing their utility in understanding and predicting system behaviour.

@@

	In their comprehensive review, \citet{Saltelli2005} examined sensitivity analysis methods applicable to chemical reaction modelling. They discussed both local and global techniques, emphasizing the role of global sensitivity analysis in understanding complex chemical systems. The authors highlighted the importance of variance-based methods in quantifying the influence of input parameters on model outputs by discussing results of a few simulation analysis.

	A key novelty of this work lies in its rigorous application of derivative-based local sensitivity analysis to a first-principles model of supercritical fluid extraction. While multiple publications available in the literature discuss application of sensitivity analysis on a supercritical extraction model, many of these either omit local sensitivity analyses and focus on basic parametric comparisons (\citet{Santos2000}, \citet{Fiori_2007}, or \citet{Zahedi2010}). In contrast, the present approach integrates a quasi-1D partial-differential-equation model with automatic differentiation of the mass and energy balances, enabling a detailed exploration of how infinitesimal changes in parameters, which propagates propagate through the system. As the result, corresponding system response of the state space (with emphasis on solutes concentrations on solid and liquid phase as well as yield). Although this work focus on the influence of a few selected parameters, the same methodology can be extended to any other parameter. Notably, this level of detail is crucial near the critical region of CO$_2$, where small deviations in process conditions can trigger large changes in solvent properties (such as density and solubility). This approach thus provides a systematic and quantitative measure of the system’s sensitivity to parameter fluctuations across different extraction regimes, from kinetic- to diffusion-controlled.

@@

	The set of quasi-one-dimensional governing equations in Cartesian coordinates is described by Equations \ref{EQ: CompressibleEuler_1}--\ref{EQ: CompressibleEuler_3}:

@@

	where $\rho_f$ is the density of the fluid, $A_f$ is the function which describes a change in the cross-section, $v$ is the velocity, $P$ is the total pressure, $e$ is the internal energy of the fluid, $t$ is time, and $z$ is the spatial direction.

@@

	Assuming that the mass flow rate is constant in time, the temporal derivative becomes the mass flux F, and the spatial derivative can be integrated along $z$ as follows:

@@

	To simplify the system dynamics, it is assumed that $F$ is a control variable and affects the whole system instantaneously (due to $\nabla \cdot u = 0$), which allows finding the velocity profile that satisfies mass continuity based on $F$, $\phi$, and $\rho_f$:

@@

	The fluid density $\rho_f$ can be obtained from the Peng--Robinson equation of state if the temperature and thermodynamic pressure are known along $z$. Variation in fluid density may occur due to pressure or inlet temperature changes. In a non-isothermal case, in Equations \ref{EQ:Velocity_1} and \ref{EQ:Velocity_2}, $\rho_f$ is considered the average fluid density along the extraction column.

@@

	As the solvent flows through the fixed bed, CO$_2$ molecules diffuse into the pores, adsorb on the inner surface, and form a film due to solvent--solid matrix interactions. The dissolved solute diffuses into the bulk from the particle's core through the solid--fluid interface, the pore and the film. Figure \ref{fig: SFE_Mechanism} shows the mass transfer mechanism, where the mean solute concentration in the solid phase is denoted by $c_s$, and the equilibrium concentrations at the solid--fluid interface are denoted by $c_s^*$ and $c_p^*$ for the solid and fluid phases, respectively. The concentration of the solutes in the fluid phase in the centre of the pore is denoted by $c_p$. As the solute diffuses through the pore, its concentration changes, reaching $c_{pf}$ at the opening. Then, the solute diffuses through the film around the particle and reaches bulk concentration $c_f$. The two-film theory describes the solid-fluid interface inside the pore. The overall mass transfer coefficient can be determined from the relationship between the solute concentration in one phase and its equilibrium concentration.

@@

	where $\mu$ is sphericity, $l$ a characteristic dimension of particles that can be defined as $l = r/3$, $r$ is the mean particle radius, $\rho_s$ is the solid density, $D_i$ corresponds to the overall diffusion coefficient, and $c_P^*$ is the concentration at the solid--fluid interface (which, according to the internal resistance model, is supposed to be at equilibrium with the fluid phase).

@@

	The volumetric partition coefficient $k_p$ acts as an equilibrium constant between the solute concentration in one phase and the corresponding equilibrium concentration at the solid--fluid interphase. \citet{Spiro2007} propose defining the mass partition coefficient $k_m$ as follows:

@@

	According to \citet{Reverchon1996}, the kinetic term becomes the following:

@@

	The choice between the two formulations should consider the system's properties and the underlying mass transfer assumptions. In systems where external mass transfer dominates, the resistance to solute transport primarily occurs at the fluid--solid interface, typically governed by the diffusion of solute across the fluid boundary layer. Under these conditions, the Bulley formulation is more appropriate, as it explicitly emphasizes the fluid-phase concentration ($c_f$) as the driving force for mass transfer through the term $c_f - c_p^*(c_s)$. Conversely, when internal mass transfer dominates, the resistance to solute transport is controlled by diffusion or adsorption processes within the solid phase. In such cases, the Reverchon formulation is more suitable, as it accounts for deviations in the equilibrium through the term $c_s - c_s^*(c_f)$.

@@

	Following the idea of the broken-and-intact cell (BIC) model of \citet{Sovova2017}, the internal diffusion coefficient $D_i$ is considered to be a product of the reference value of $D_i^R$ and the exponential decay function $\gamma$, as given by Equation \ref{EQ: C_sat_function}:

@@

	The empirical correlations for $D_i$ and $\Upsilon$ were derived by \citet{Sliczniuk2024} and validated for temperatures of $30 - 40^\circ C$, pressures of $100 - 200$ bar, and mass flow rates of $3.33-6.67 \cdot 10^{-5}$ kg/s. Figures \ref{fig:Correlation_Di} and \ref{fig:Correlation_Gamma} show the results of multiple linear regression applied to solutions of parameter estimation and selected independent variables. The region marked with the white dashed line represents the confidence region, where the model has been tested. Both correlations should be equal or greater than zero to avoid unphysical behaviour such as  reverse mass transfer. The multiple linear regression functions are combined with the rectifier function to ensure non-negativity.

@@

	If the value of enthalpy $h$ is known from the time evolution of the energy equation, and pressure $P$ is known from measurement, then temperature $T$ can be reconstructed based on the departure function. The departure function is a mathematical function that characterizes the deviation of a thermodynamic property (enthalpy, entropy, or internal energy) of a real substance from that of an ideal gas at the same temperature and pressure. The enthalpy departure function, as presented by \citet{Gmehling2019} for the Peng--Robinson equation of state,  is defined by Equation \ref{EQ:Enthalpy_PR}:

@@

	where $\alpha$ is defined as $\left( 1+\kappa \left( 1 - \sqrt{T_r} \right) \right)^2$, $T_r$ is the reduced temperature, $P_r$ is the reduced pressure, $Z$ is the compressibility factor, $\kappa$ is a quadratic function of the acentric factor, and $B$ is calculated as $0.07780\frac{P_r}{T_r}$.

@@

	The method of lines is used to transform the process model equations into a set of ODEs denoted by $G(x;\Theta)$. For a derivative to be conservative, it must form a telescoping series. In other words, only the boundary terms should remain after adding all terms coming from the discretization over a grid, and the artificial interior points should be cancelled out. Discretization is applied to the conservative form of the model to ensure mass conservation. The backward finite difference is used to approximate the first-order derivative, while the central difference scheme approximates the second-order derivative $z$ direction. The length of the fixed bed is divided into $N_z$, that is, equally distributed points in the $z$ direction. The state-space model after discretization is denoted by $x$ and defined as follows the Figure \ref{fig:discretization}, where $x \in \mathbb{R}^{N_x = 3N_z+2} $ and $\Theta \in \mathbb{R}^{N_\Theta =  N_{\theta} + N_u } $, $N_{\theta}$ is the number of parameters and $N_{u}$ is the number of control variables.

@@

	The sensitivity matrix $S(x;\Theta)$ represents how the state vector $x$ changes in response to variations in the model parameters $\Theta$. In physical terms, this means it measures how sensitive the system is to changes in the underlying parameters that define its behaviour. The new system of equations can be obtained by taking the total derivative of $S$ with respect to time $t$:

@@

	This equation decomposes the total effect of parameter changes on the system dynamics into two parts. The indirect effect is represented by the term involving $\bar{J}_x (x; \Theta)$. This term shows how changes in the parameters influence the system through their effect on the state variables. This is an indirect feedback effect where parameter changes alter the state, which in turn affects the system response. The direct effect given by $\bar{J}_x (x; \Theta)$ reflects the immediate sensitivity of the system to parameter variations.

@@

	The augmented system combines both the state equations and sensitivity equations into a unified system. This allows the simultaneous computation of both the system behaviour and its sensitivity to parameter changes.

	The initial conditions are described as follows:

@@

	Details of the process model and its parameters can be found in the work of \citet{Sliczniuk2024}. The model was validated under the following operating conditions: temperatures from $30$ to $40 ^\circ C$, pressures from $100$ to $200$ bar, and mass flow rates from $3.33$ to $6.67 \cdot 10^{-5}$ kg/s. In this study, the local sensitivity analysis is used to investigate how changes in model parameters affect the state space and extraction yield, where the results reflect the system’s response to infinitesimal deviations in parameter values. Although this work focuses on two specific cases, the methodology can be extended to any model parameter. The first case assesses the empirical correlations embedded in the process model, specifically the zeroth-order term $\Upsilon(0)$ in the multiple linear regression for $\Upsilon$. The second case examines the effect of operating conditions, particularly pressure, on the state space and model outputs. For the first analysis, the midpoint of the validated range ($35 ^\circ C$, $150$ bar, and $5\cdot 10^{-5}$) was selected. The second analysis is carried out in two steps. First, we examine the impact of pressure changes on the state variables (solute concentrations in both solid and fluid phases) at the same midpoint conditions. Second, we perform a yield analysis at the same temperature and mass flow rate, but at different pressures (100, 125, 150, 175, and 200 bar).
